Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-04 manuscript.tex
latest

The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated. Let $k$ denote the total number of propagules in a territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$. Then, applying Bayes' theorem, $\tilde{p}$ can be written as
\begin{align}
\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|k\geq 2,x_i=1),\nonumber\\
&=\frac{P({\mathbf x_i},k\geq 2|x_i=1)}{P(k\geq 2)},\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|\sum_{j\neq i} x_j=k-1),\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} \frac{P(k)\delta^{\sum_{j\neq i} x_j}_{k-1}}{P(\sum_{j\neq i} x_j=k-1)} p({\mathbf x_i}),
\label{eq:raremonster1}
\end{align}
where $P(k)=L^k e^{-L}/k!$, $P(\sum_{j\neq i} x_j=k-1)=(L-l_i)^{k-1} e^{-(L-l_i)}/(k-1)!$ and $\delta^{\sum_{j\neq i} x_j}_k=1$ if $\sum_{j\neq i} x_j=k$, and equals zero otherwise. This implies $P(k)/P(\sum_{j\neq i} x_j=k-1)=\frac{Le^{-l_i}}{k}\left(\frac{L}{L-l_i}\right)^{k-1}$, and so, changing the summation index to start from one, we obtain
\begin{equation}
\tilde{p}({\mathbf x_i})=\frac{Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{\delta^{\sum_{j\neq i} x_j}_k}{k+1} p({\mathbf x_i}).
\label{eq:raremonster2}
\end{equation}

To evaluate $\langle x_j \rangle_{\tilde{p}}=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j$, we first evaluate the sum $\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j$, since the only term in Eq. \eqref{eq:raremonster2} which depends on ${\mathbf x_i}$ is $\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})$. By inspection, this sum is equal to the mean number of $j$ propagules in a territory which received $k$ nonfocal propagules, times the probability that a territory receives $k$ nonfocal propagules. Given that $k$ nonfocal propagules are present, on average genotype $j$ will make up $\frac{l_j}{L-l_j}k$ of them, and so
\begin{align}
\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j&= \frac{l_j}{L-l_j}k \frac{(L-l_i)^{k} e^{-(L-l_i)}}{k!}\nonumber\\
&=l_j \frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k-1)!}.
\end{align}
The final step is to evaluate the sum over $k$, 
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k+1)(k-1)!}.\nonumber\\
&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^k e^{-(L-l_i)}}{(k+1)!}k.\nonumber\\
&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k.\nonumber\\
&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
\label{eq:meanxjrare}
\end{align}
where the last line follows from the fact that $\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k=\sum_{k=1}^{\infty} \frac{L^{k} e^{-L}}{k!}(k-1)=L-(1-e^{-L})$.

For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq. \eqref{eq:meanxjrare} is unnecessarily complicated. We instead use the following approximation. Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $\sum_{j\neq i} x_j>1$, in a territory where one focal propagule is assumed to already be present. This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$, 
\begin{equation}
\sigma_{\tilde{q}}^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
\end{equation}
and 
\begin{equation}
\sigma_{\tilde{q}}(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
\end{equation}
where $C=1-e^{-(L-l_i)}$. Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, the subtle difference in how we defined $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$). To be sure, $\tilde{q}$ is incorrect: assuming that one propagule is already present biases total number of propagules by one. For example, if the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$), Eq. \eqref{eq:meanxjrare} correctly gives $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\langle x_j \rangle_{\tilde{q}}\approx L$. As a result, $\tilde{q}$ leads to pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to $\tilde{p}$'s that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.

>>>>

The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated. Let $X$ denote the total number of propagules in a territory,  ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$, and let $X_i=X-x_i$ denote the total number of nonfocal propagules. Then, $\tilde{p}$ can be written as
\begin{align}
\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|X\geq 2,x_i=1)\nonumber\\
&=\frac{P({\mathbf x_i},X\geq 2|x_i=1)}{P(X\geq 2)}\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|X_i=k-1),
\end{align}
and so
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j,\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=k-1)x_j.
\label{eq:raremonster1}
\end{align}
The inner sum over ${\mathbf x_i}$ is the mean number of $j$ propagules that will be found in a territory which received $k-1$ nonfocal propagules, which is equal to $\frac{l_j}{L-l_j}(k-1)$. Thus, 
\begin{align}
\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(k) (k-1),\nonumber\\
&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
\label{eq:meanxjrare}
\end{align}
where the last line follows from the fact that $\sum_{k=2}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)k-\sum_{k=1}^{\infty}P(k)=L-(1-e^{-L})$.

For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq. \eqref{eq:meanxjrare} is unnecessarily complicated, and so we instead use the following approximation. Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $X_i>1$, in a territory where one focal propagule is assumed to already be present. This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$, 
\begin{equation}
\sigma_{\tilde{q}}^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
\end{equation}
and 
\begin{equation}
\sigma_{\tilde{q}}(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
\end{equation}
where $C=1-e^{-(L-l_i)}$. Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, the subtle difference in how we defined $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$). To be sure, $\tilde{q}$ is incorrect: assuming that one propagule is already present biases total number of propagules by one. For example, if the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$), Eq. \eqref{eq:meanxjrare} correctly gives $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\langle x_j \rangle_{\tilde{q}}\approx L$. As a result, $\tilde{q}$ leads to pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to $\tilde{p}$'s that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
